Network Visualizations in \MakeLowercase{\texttt{ggplot2}}

### 2015-10-15 inst/paper.Rnw
default arrow size set in point
bug fix

\subsection{geom\_net}
\subsubsection{Data Structure}
\hh{Network analysis is usally working with two sources of information: one data set consisting of a description of the `players', represented as the nodes or vertices in the network, and another data set detailing the relationship between the players, the edge dataset.
In order for this geometry to work, these two data sets need to be combined into a single dataset. For this, we are using the convention that all of the node information is merged into the edge data set using the `from' as the reference column. Generally, there will be some vertices that are sinks in the network, ie.\ they only show up in the `to' column. We can easily accommodate for these nodes by adding artificial edges in the data set, that have missing information for the `to' column.
Fortunately, R provides functions that allow for an easy way of producing the required result: both \code{merge} and \code{join} can be used. In \code{merge} the paramter all needs to be set to TRUE, in \code{join}, the parameter setting has to be \code{type=`full'}, and \code{all=TRUE}.

@@

\caption{\label{fig:email_ex_facet} The same email network as in figure~\ref{fig.cap:email_ex} facetted by day of the week.}
\end{figure}
%\end{landscape}
This plot is shown in Figure~\ref{fig:email_ex_facet}. With the facetting, we can see that there are several days where one or more departments do not communicate with any of the other departments. There are only two days, 13 and 15, without any isolated department communications. Facetting is one of the major benefits of creating a geometry for networks in \texttt{ggplot2}.  Facetting quickly separates dense networks into separate subnetworks for easy visual comparison and  analyses.
  \subsection{\texttt{ggplot2} Theme Elements}
This example comes from the \texttt{theme()} help page in the \texttt{ggplot2} documentation \citep{ggplot}.  It is a directed network which shows the structure of the inheritance of theme options in the construction of a \texttt{ggplot2} plot.  There are 53 vertices and 36 edges in this network. Each vertex represents one possible theme option. There is an arrow from one theme option to another if the element represented by the \emph{to} vertex inherits its values from the \emph{from} vertex.  For example, the \texttt{axis.ticks.x} option inherits its value from the \texttt{axis.ticks} value, which in turn inherits its value from the \texttt{line} option.  Thus, setting the \texttt{line} option to a value such as \texttt{element\_blank()} sets the entire inheritance tree to \texttt{element\_blank()}, and no lines appear anywhere on the plot background. Finally, we note that the vertices with no edges were incorporated into the plot by adding their labels to the edges data frame in both the  `from\_id' and `to\_id' columns before passing the edges data frame to \texttt{ggplot}.
<<theme_ex, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
ggplot(data = TEnet, aes(from_id = parent, to_id = child)) +
  geom_net(aes(fontsize=degree), directed = TRUE, label = TRUE, vjust=-.5,
           size=3, ecolour="grey70") +
  theme_net() + xlim(c(-0.05, 1.05))
@
<<theme_ex_ggnet2, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
te.net = network(theme_elements$edges)
te.net %v% "size" = sqrt(10 * (sna::degree(te.net) + 1))
ggnet2(te.net, label = TRUE, color = "white",
       label.size = "size", layout.exp = 0.15)
@
<<theme_ex_ggnetwork, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
ggplot(ggnetwork(te.net), aes(x, y, xend = xend, yend = yend)) +
  geom_edges() +
  geom_nodes(size = 12, color = "white") +
  geom_nodetext(aes(size = size, label = vertex.names)) +
  scale_size_continuous(range = c(3, 5)) +
  guides(size = FALSE) +
  theme_blank()
@

>>>>

\subsection{geom\_net}
\subsubsection{Data Structure}
\hh{Network analysis is usally working with two sources of information: one data set consisting of a description of the `players', represented as the nodes or vertices in the network, and another data set detailing the relationship between the players, the edge dataset.
In order for this geometry to work, these two data sets need to be combined into a single dataset. For this, we are using the convention that all of the node information is merged into the edge data set using the `from' as the reference column. Generally, there will be some vertices that are sinks in the network, ie.\ they only show up in the `to' column. We can easily accommodate for these nodes by adding artificial edges in the data set, that have missing information for the `to' column.
Fortunately, R provides function that allow for an easy way of producing the required result: both \code{merge} and \code{join} can be used. In \code{merge} the paramter all needs to be set to TRUE, in \code{join}, the parameter setting has to be \code{type=`full'}, and \code{all=TRUE}.

@@

\caption{\label{fig:email_ex_facet} The same email network as in figure~\ref{fig.cap:email_ex} facetted by day of the week.}
\end{figure}
%\end{landscape}
This plot is shown in Figure~\ref{fig:email_ex_facet}. With the facetting, we can see that there are several days where one or more departments do not communicate with any of the other departments. There are only two days, 13 and 15, without any isolated department communications. Facetting is one of the major benefits of creating a geometry for networks in \texttt{ggplot2}.  Facetting quickly separates dense networks into separate subnetworks for easy visual comparison and  analyses.
  \subsection{\texttt{ggplot2} Theme Elements}
This example comes from the \texttt{theme()} help page in the \texttt{ggplot2} documentation \citep{ggplot}.  It is a directed network which shows the structure of the inheritance of theme options in the construction of a \texttt{ggplot2} plot.  There are 53 vertices and 36 edges in this network. Each vertex represents one possible theme option. There is an arrow from one theme option to another if the element represented by the \emph{to} vertex inherits its values from the \emph{from} vertex.  For example, the \texttt{axis.ticks.x} option inherits its value from the \texttt{axis.ticks} value, which in turn inherits its value from the \texttt{line} option.  Thus, setting the \texttt{line} option to a value such as \texttt{element\_blank()} sets the entire inheritance tree to \texttt{element\_blank()}, and no lines appear anywhere on the plot background. Finally, we note that the vertices with no edges were incorporated into the plot by adding their labels to the edges data frame in both the  `from\_id' and `to\_id' columns before passing the edges data frame to \texttt{ggplot}.
<<theme_ex, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
ggplot(data = TEnet, aes(from_id = parent, to_id = child)) +
  geom_net(aes(fontsize=degree), directed = TRUE, label = TRUE, vjust=-.5,
           size=3, ecolour="grey70", curvature=0.25) +
  theme_net() + xlim(c(-0.05, 1.05))
@
<<theme_ex_ggnet2, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
te.net = network(theme_elements$edges)
te.net %v% "size" = sqrt(10 * (sna::degree(te.net) + 1))
ggnet2(te.net, label = TRUE, color = "white",
       label.size = "size", layout.exp = 0.15)
@
<<theme_ex_ggnetwork, echo=TRUE, fig.cap= 'Inheritance structure for \\texttt{theme} elements in \\texttt{ggplot2}.'>>=
ggplot(ggnetwork(te.net), aes(x, y, xend = xend, yend = yend)) +
  geom_edges() +
  geom_nodes(size = 12, color = "white") +
  geom_nodetext(aes(size = size, label = vertex.names)) +
  scale_size_continuous(range = c(3, 5)) +
  guides(size = FALSE) +
  theme_blank()
@
